AMBIENT; an execution and an MPI-CUDA scheduler

### 2011-08-12 doc/paper_ambient/paper_ambient.tex
live session corrections

When the framework encounters the signal to flush the stack of kernels the actual execution starts. The framework's engine goes through the accumulated stack 
and executes all independent operations at once. Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
amount of resources for this operation. Next, the engine marks argument's tiling using specified parameters (???) and maps resulting work groups onto allocated
processes in a user specified way. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA, with work groups
simply being mapped onto threads of MPI process as if they were GPU SMP (the difference lies mainly in the user's control over the mapping of processes to work groups).

@@

The work group approach imposes the adaptation of the data layout to the framework. In linear algebra,  BLAS \cite{LAWSON-1979} is \textit{de facto} the standard for high performance dense Linear Algebra computations. It imposes column order on matrix storage, as well as a contiguous storage in memory (figure \ref{MEMMATRIX}).

For the sake of backward compatibility with the standard BLAS-based mathematical libraries (BLAS, LaPACK ScaLAPACK), we keep, the column order, but do not conserve a full contiguous buffer. As in the plasma library \cite{LTAIEF-2011}, we select
a tile data layout. It confers the ability to fit perfectly with the work group representation of Ambient, and further allows an excellent MPI/Thread control (section \ref{MPIABSTRACTION}). One of the main features of Ambient is to map this new data layout  to our work group approach, as shown in Fig. \ref{CUDAANALOGY}.

@@

In linear algebra, the optimal distribution in terms of load balancing is the 2D block-cyclic distributions, esp. in ScaLAPACK, but this will generally depend of the solver used. For example, in a QR factorization \cite{DEMMEL-2008} it can be demonstrated that a row-wise distribution is superior to a 2D block-cyclic decomposition for matrices with more rows than columns. Switching from one distribution scheme to another, which is a challenge in ScaLAPACK, is easily accomplished with Ambient.
 \begin{table}[h]
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           \textbf{Logistic kernel }  Algorithm of the addition logistic kernel  &    \\ \hline
                           \textbf{Require}                 A class \texttt{Matrix}  that fits with the Ambient framework  &  \\
                           \begin{tabular}{c l} 
                           \tiny{1:} &  Addition\_logistic(Matrix $A$,  Matrix $B$)  \\
                           \tiny{2:} &  SQL request for selecting $N$ number of processors   \\
                           \tiny{3:} &  Perform 2/1-D cyclic distribution of the Matrix $A$  $\vartriangleright$ Note: the distribution \\
                                         & is performed with the number of processors selected inside the SQL request. \\
                           \tiny{4:} &  Perform 2/1-D cyclic distribution of the Matrix $B$ \\
%                           \tiny{5:} &  Perform 2/1-D cyclic distribution of the Matrix $C$ \\ 
                           \end{tabular} & \\ \hline
		 \end{tabular}  
		 \caption{Logistic kernel associated to the matrix multiplication or  the addition.  \label{LKERNEL}}
	\end{center}
\end{table}

To illustrate, we describe the example of a simplified\footnote{without any C++ language annotations} logistic kernel of the matrix addition in table {\ref{LKERNEL}}. 
\begin{itemize}
\item Line 1:  the signature of the logistic kernel, showing the input/ouput by the keyword\texttt{const} (???). In this example, there are only two matrices, but additional arguments like float numbers, integers \etc can be added.
\item Line 2: The SQL-like request selecting the desired number of processors for the operation and performing the required MPI calls. 
\item Line 3-5: Performing a block-cyclic distribution of the matrixes \texttt{A} and  \texttt{B}. A detailed description of the SQL requests and the distribution functions
is given in section \ref{SQLSECTION}. Note that the present kernel can be reused for any subsequent operation that uses the same data distribution.
\end{itemize}

@@

We have introduced a keyword (as the \texttt{const} keyword (???)) named \texttt{pinned}. This keyword indicates to the framework that the kernel must be executed over all Ambient work groups. Inside a kernel, we employ several Ambient API, analogous to CUDA, to obtain key information about the work groups 
(positions, dimensions, \ldots).These are summarized in table \ref{ANALOGY}.  For illustration, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with the previous example of the logistic kernel is presented in the table \ref{LKERNEL}.
\begin{itemize}
\item Line 1: the signature of the computational kernel must be the same than that of the associated logistic kernel. The only difference is the use of the keyword \texttt{pinned}, indicating an execution over all Ambient work groups.
\item Line 2: the cartesian coordinates of the work group are obtained from the Ambient instruction \texttt{get\_block\_id(A).x}  for $i$ and \texttt{get\_block\_id(A).y}  for $j$. \label{CURRENT}
\item Line 3 to 5 : we refer the work group $i,j$ to a temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}. Note that if the processor requires data outside its memory,
 it will perform a remote access in the background (see section \ref{MPIABSTRACTION} for details).
\item Line 6 to 8 : the addition is performed by looping over the total size of the work group, in parallel for each workgroup (because of the use of the keyword \texttt{pinned}).
\end{itemize}

@@

As just shown for the matrix addition, we can recode other linear algebra solvers with the Ambient approach. For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as a possible alternative to the pDGEMM solver, called (???), using a recursive algorithm.

@@

For the packet manager to be adaptable to any communication pattern, it consists of two parts:
the delegate pattern and a finite state machine (FSM), both are placed inside this loop.

@@

The delegate pattern (from C\#) encapsulates  methods, behaving similar to a function pointer in C++,  but allowing for safer execution.
The delegate object inside the packet manager class contains a series of functions piled up in a stack, such as forwarding requests, 
layout, and block of data (???). As this delegate is inside the infinite loop, the delegate execution 
will propagate the packets process by process after every iteration (as shown in Fig. \ref{EXECUTION}). When the delegate execution is over, a finite state machine
checks the states of the packets (arrived or not).  If all packets (data packet, layout packet, etc.) have been received, the FSM breaks the infinite loop.

>>>>

When the framework encounters the signal to flush the stack of kernels the actual execution starts. The framework's engine goes through the accumulated stack 
and executes all independent operations at once. Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
amount of resources for this operation. Next, the engine sets the argument's (here: the matrices) tiling using user specified work group sizes and maps resulting work groups onto allocated
processes. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA, with work groups
simply being mapped onto threads of MPI process as if they were GPU SMP (the difference lies mainly in the user's control over the mapping of processes to work groups).

@@

The work group approach imposes the adaptation of the data layout to the framework. The standard math libraries impose both column order as well as a contiguous memory order on matrix storage (figure \ref{MEMMATRIX}).

Therefore, we keep the column order but do not conserve a full contiguous buffer. As in the PLASMA library \cite{LTAIEF-2011}, we choose
a tile data layout. It confers the ability to fit perfectly with the work group representation of Ambient, and further allows an excellent MPI/Thread control (section \ref{MPIABSTRACTION}). One of the main features of Ambient is to map this new data layout  to our work group approach, as shown in Fig. \ref{CUDAANALOGY}.

@@

In linear algebra, the optimal distribution in terms of load balancing is the 2D block-cyclic distributions, esp. in ScaLAPACK, but this will generally depend of the solver used. For example, in a QR factorization \cite{DEMMEL-2008} it can be demonstrated that a row-wise distribution is superior to a 2D block-cyclic decomposition for matrices with more rows than columns. Switching from one distribution scheme to another, which is a challenge in ScaLAPACK, is easily accomplished with Ambient.
 \begin{table}[h]
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           \textbf{Logistic kernel }  Algorithm of the addition logistic kernel  &    \\ \hline
                           \textbf{Require}                 A class \texttt{Matrix}  that fits with the Ambient framework  &  \\
                           \begin{tabular}{c l} 
                           \tiny{1:} &  Addition\_logistic(Matrix $A$,  const Matrix $B$)  \\
                           \tiny{2:} &  SQL request for selecting $N$ number of processors   \\
                           \tiny{3:} &  Perform 2/1-D cyclic distribution of the Matrix $A$  $\vartriangleright$ Note: the distribution \\
                                         & is performed with the number of processors selected inside the SQL request. \\
                           \tiny{4:} &  Perform 2/1-D cyclic distribution of the Matrix $B$ \\
%                           \tiny{5:} &  Perform 2/1-D cyclic distribution of the Matrix $C$ \\ 
                           \end{tabular} & \\ \hline
		 \end{tabular}  
		 \caption{Logistic kernel associated to the matrix multiplication or  the addition.  \label{LKERNEL}}
	\end{center}
\end{table}

To illustrate, we describe the example of a simplified\footnote{without any C++ language annotations} logistic kernel of the matrix addition in table {\ref{LKERNEL}}. 
\begin{itemize}
\item Line 1:  the signature of the logistic kernel.
\item Line 2: The SQL-like request selecting the desired number of processors for the operation and performing the required MPI calls. 
\item Line 3-5: Performing a block-cyclic distribution of the matrixes \texttt{A} and  \texttt{B}. A detailed description of the SQL requests and the distribution functions
is given in section \ref{SQLSECTION}. Note that the present kernel can be reused for any subsequent operation that uses the same data distribution.
\end{itemize}

@@

We have introduced a keyword (like the \texttt{const} keyword) named \texttt{pinned}. This keyword indicates to the framework that the kernel must be executed over all work groups of the marked argument. Inside a kernel, we employ several Ambient API, analogous to CUDA, to obtain key information about the work groups 
(positions, dimensions, \ldots).These are summarized in table \ref{ANALOGY}.  For illustration, the computational kernel (table \ref{CKERNEL}) of an addition of the two matrices associated with the previous example of the logistic kernel is presented in the table \ref{LKERNEL}.
\begin{itemize}
\item Line 1: the signature of the computational kernel must be the same than that of the associated logistic kernel. The only difference is the use of the keyword \texttt{pinned}, indicating an execution over all Ambient work groups.
\item Line 2: the cartesian coordinates of the work group are obtained from the Ambient instruction \texttt{get\_block\_id(A).x}  for $i$ and \texttt{get\_block\_id(A).y}  for $j$. \label{CURRENT}
\item Line 3 to 5 : we refer the work group $i,j$ to a temporary variable $A_{ref}$ using the Ambient instruction \texttt{current(A)(i,j)}. Note that if the processor requires data outside its memory,
 it will perform a remote access in the background (see section \ref{MPIABSTRACTION} for details).
\item Line 6 to 8 : the addition is performed by looping over the total size of the work group, in parallel for each workgroup (because of the use of the keyword \texttt{pinned}).
\end{itemize}

@@

As just shown for the matrix addition, we can recode other linear algebra solvers with the Ambient approach. For the presentation of our framework, we will focus on the MPI version of BLAS named pBLAS \cite{SCALAPACK-1992}, as well as recursive Strassen-like recursive reduced multiplication algorithms.

@@

For the packet manager to be adaptable to any communication pattern, it employs two parts:
a delegate mechanism and a finite state machine (FSM).

@@

The delegate mechanism serves its design goal - to execute handlers of components listening for the packets arrival, such as layout tables.
This delegate object inside the packet manager is tied to a list of functions, such as layout forwarding, 
data block forwarding and control information handlers. As the packet manager checks for completion of the local send process, the finite state machine
checks the send completion state of the other MPI processes (whether they have sent all their packets or not). If all the packets have been sent, the FSM breaks the infinite loop.
